---
title: "Heart Disease Patient Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

Heart disease is one of the leading causes of death worldwide, and identifying
patterns in patient health data can help in early detection and prevention.

This project analyzes the Heart Disease dataset (originally from the UCI
Machine Learning Repository and widely used on Kaggle as the "Heart Disease
UCI" dataset) using R programming. The analysis focuses on patient
characteristics such as chest pain type, thalassemia test results, cholesterol
levels, and heart disease diagnosis outcomes.

The project uses statistical analysis and data visualization to identify
important patterns and trends related to heart disease in the dataset.

# 2. Objectives

The main objectives of this project are:

To analyze patient records related to heart disease.
To study cholesterol level distributions.
To identify common chest pain types among patients.
To analyze the distribution of thalassemia test results among patients.
To examine the relationship between age and maximum heart rate.
To visualize important patterns using graphs.
To interpret the results and present meaningful findings.

# 3. Data Importing

```{r}
heart_data <- read.csv("heart_disease_dataset.csv")

head(heart_data)
```

# 4. Understanding the Dataset

```{r}
str(heart_data)

summary(heart_data)
```

# 5. Checking Missing Values

```{r}
colSums(is.na(heart_data))
```

# 6. Data Cleaning

The dataset uses numeric codes for some categorical variables. These are
converted into readable labels to make the analysis and charts easier to
interpret.

```{r}
heart_data$cp_label <- factor(
  heart_data$cp,
  levels = c(1, 2, 3, 4),
  labels = c("Typical Angina", "Atypical Angina",
             "Non-Anginal Pain", "Asymptomatic")
)

heart_data$sex_label <- factor(
  heart_data$sex,
  levels = c(0, 1),
  labels = c("Female", "Male")
)

heart_data$target_label <- factor(
  heart_data$target,
  levels = c(0, 1),
  labels = c("No Disease", "Disease Present")
)

head(
  heart_data[, c("cp_label", "sex_label", "target_label")]
)
```

>The dataset contains no missing values, so no rows needed to be removed
during cleaning.

# 7. Cholesterol Analysis

>Cholesterol statistics are calculated using the cholesterol level (mg/dl)
reported for each patient.

```{r}
mean(heart_data$chol, na.rm = TRUE)
```
```{r}
median(heart_data$chol, na.rm = TRUE)
```
```{r}
var(heart_data$chol, na.rm = TRUE)
```
```{r}
sd(heart_data$chol, na.rm = TRUE)
```

# Interpretation

The mean represents the average cholesterol level across all patients in the
dataset, while the median represents the middle cholesterol value. The
difference between the mean and median indicates that the cholesterol
distribution is affected by a few unusually high observations.

# 8. Chest Pain Type Analysis

```{r}
cp_count <- table(heart_data$cp_label)

sort(cp_count, decreasing = TRUE)
```

# 9. Chest Pain Type Visualization

```{r}
barplot(
  sort(cp_count, decreasing = TRUE),
  main = "Chest Pain Type Frequency",
  xlab = "Chest Pain Type",
  ylab = "Number of Patients",
  las = 2
)
```

# Interpretation

The bar chart shows the number of patients reporting each chest pain type.
Asymptomatic chest pain is the most frequently recorded type in the dataset,
followed by non-anginal pain, atypical angina and typical angina. This helps
identify which chest pain presentations are most common among the patients
in this dataset.

# 10. Thalassemia Test Result Analysis

```{r}
thal_count <- table(heart_data$thal)

thal_count
```
```{r}
barplot(
  thal_count,
  main = "Thalassemia Test Result Distribution",
  xlab = "Thalassemia Result",
  ylab = "Number of Patients",
  las = 2
)
```

# Interpretation

The bar chart shows the distribution of thalassemia test results across the
patients in the dataset. A "normal" result is the most common outcome,
followed by "reversible" and "fixed" defects. This shows that most patients
in the dataset returned a normal thalassemia test result.

# 11. Heart Disease Diagnosis Analysis

```{r}
target_count <- table(heart_data$target_label)

target_count
```
```{r}
pie(
  target_count,
  main = "Heart Disease Presence vs Absence",
  labels = paste(names(target_count), target_count)
)
```

# Interpretation

The pie chart compares patients diagnosed with heart disease against those
without it. Out of the total patients in the dataset, a smaller portion is
identified as having heart disease, while the larger portion does not have
the condition.

# 12. Cholesterol Distribution

```{r}
hist(
  heart_data$chol,
  main = "Cholesterol Level Distribution",
  xlab = "Cholesterol (mg/dl)",
  ylab = "Frequency"
)
```

# Interpretation

The histogram shows the distribution of cholesterol levels reported for the
patients. Most observations are concentrated in a moderate cholesterol range,
while a small number of patients have considerably higher cholesterol values.
This indicates that the cholesterol distribution is right-skewed and contains
some high-value observations.

# 13. Age vs Maximum Heart Rate Analysis

```{r}
plot(
  heart_data$age,
  heart_data$thalach,
  main = "Age vs Maximum Heart Rate Achieved",
  xlab = "Age",
  ylab = "Maximum Heart Rate (thalach)",
  pch = 19
)
```

# Interpretation

The scatter plot shows the relationship between a patient's age and the
maximum heart rate achieved during testing. The points show a general
downward pattern, indicating that older patients tend to achieve a lower
maximum heart rate than younger patients.

# 14. Correlation Analysis

```{r}
cor(
  heart_data$age,
  heart_data$thalach,
  use = "complete.obs"
)
```

# Interpretation

The correlation coefficient between age and maximum heart rate is
approximately -0.41. This indicates a moderate negative relationship between
the two variables.

This means that as a patient's age increases, their maximum heart rate tends
to decrease. However, correlation does not imply causation.

# 15. Findings

Based on the analysis of the Heart Disease dataset, the following findings
were obtained:

1) The dataset contains patient records with no missing values across any
of the 14 variables.

2) Asymptomatic chest pain is the most frequently reported chest pain type
among patients, followed by non-anginal pain, atypical angina and typical
angina.

3) A "normal" thalassemia test result is the most common outcome among
patients, followed by "reversible" and "fixed" defects.

4) A smaller portion of patients in the dataset are diagnosed with heart
disease compared with those who are not.

5) The cholesterol distribution shows that most patients have moderate
cholesterol levels, with a small number of unusually high observations.

6) The correlation coefficient between age and maximum heart rate is
approximately -0.41, indicating a moderate negative relationship between
the two variables.

7) The visualizations provide useful insights into chest pain types,
thalassemia results and cholesterol patterns among heart disease patients.

# 16. Conclusion

This project analyzed a real-world Heart Disease dataset using R programming.

The analysis included data importing, data exploration, missing-value
analysis, data cleaning, statistical analysis and data visualization.

The study examined chest pain types, thalassemia test results, cholesterol
levels and the relationship between age and maximum heart rate among
patients. It also compared the presence and absence of heart disease across
the dataset.

The correlation analysis showed a moderate negative relationship between age
and maximum heart rate, with a correlation coefficient of approximately
-0.41.

Overall, this project demonstrates how R can be used to analyze real-world
health data and convert it into meaningful statistical insights and
visualization.

# 17. References

Heart Disease Dataset (UCI Machine Learning Repository / Kaggle)
R Programming Documentation
R Markdown Documentation
